Fluid-mechanical erosion creates anisotropic porous media

### 2022-05-08 docs/PhysicsPaper/StokesErosion.tex
NM

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% COMMANDS
% Basic editing
\newcommand{\tocite}{{\color{blue}(to cite)}}
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
% For real and imaginary, could use \Re or \Im, or \mathcal{R}, or \text{Re}
\newcommand{\Real}{\Re}
\newcommand{\Imag}{\Im}

%---------------------------------------------------%
% From JFM 2020
\newcommand{\bd}{{\partial}}
\newcommand{\bigO}{{\mathcal{O}}}
\newcommand{\cc}{{\mathbf{c}}}
\newcommand{\CC}{{\mathbb{C}}}
\newcommand{\DD}{{\mathcal{D}}}
\newcommand{\DDD}{{\boldsymbol{\mathcal D}}}
\newcommand{\eeta}{{\boldsymbol\eta}}
\newcommand{\ff}{{\mathbf{f}}}
\newcommand{\grad}{{\nabla}}
\newcommand{\II}{{\mathbf{I}}}
\newcommand{\iin}{\mathrm{in}}
\newcommand{\llambda}{{\boldsymbol\lambda}}
\newcommand{\nn}{{\mathbf{n}}}
\newcommand{\NN}{{\mathcal{N}}}
\newcommand{\out}{\mathrm{out}}
\newcommand{\rr}{{\mathbf{r}}}
\renewcommand{\Re}{{\operatorname{Re}}}
\renewcommand{\Im}{{\operatorname{Im}}}
\newcommand{\RR}{{\mathbb{R}}}
\renewcommand{\ss}{{\mathbf{s}}}
\newcommand{\ssigma}{{\boldsymbol\sigma}}
\newcommand{\tar}{\mathrm{tar}}
\newcommand{\bary}{\mathrm{bary}}
\newcommand{\trap}{\mathrm{trap}}
\newcommand{\uu}{{\mathbf{u}}}
\newcommand{\UU}{{\mathbf{U}}}
\newcommand{\vv}{{\mathbf{v}}}
\newcommand{\xx}{{\mathbf{x}}}
\newcommand{\xxi}{{\boldsymbol{\xi}}}
\newcommand{\yy}{{\mathbf{y}}}
\newcommand{\mcaption}[2]{\caption{\small \em #1}\label{#2}} 
\newcommand{\secref}[1]{\ref{#1}}
\def\gap{\hspace*{.2in}}
% Nick's below
\newcommand{\pderiv}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\ppd}[2]{\frac{\partial^2 #1}{{\partial #2}^2}}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\Vn}{V_\nn}
\newcommand{\Vs}{V_\ss}
\newcommand{\CE}{C_E}
%---------------------------------------------------%

% From Nick's Latex Notes
\newcommand{\bvec}[1]{\ensuremath{\boldsymbol{#1}}}
\newcommand {\bq} {\bvec{q}}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\qavg}{\bar{q}}
\newcommand{\pavg}{\bar{p}}
\newcommand{\pup}{p_u}
\newcommand{\pdn}{p_d}
\newcommand {\bu} {\bvec{u}}
\newcommand{\nhat}{\hat{\bvec{n}}}
\newcommand{\shat}{\hat{\bvec{s}}}
\newcommand{\FD}{\bvec{F_D}}
\newcommand{\ex}{\hat{\bvec{e}}_1}

@@

% NOTE: THIS SECTION HAS BEEN PLAGIARIZED FROM JFM 2020
% SO WE NEED TO PARAPHRASE IT
%\begin{comment}
%-----------------------------------------------------------------------------------------------%
\section{Governing Equations}
\label{sec:formulation}
%-----------------------------------------------------------------------------------------------%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\citep{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of the boundary of $[-3,3] \times [-1,1]$.  All eroding
bodies are placed in $[-1,1] \times [-1,1]$ to create a buffer region
that allows the flow profile imposed at the inlet to transition to the
more complex flow intervening between the bodies. Neglecting inertial
forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, and $\Vn$ is the normal
velocity of $\gamma$. Since the rate of erosion is much slower than the
background flow, a quasi-steady approximation justifies the no-slip
boundary condition. Parameters include the fluid viscosity $\mu$ and the
material-dependent erosion constant $\CE$. By non-dimensionalizing
equations~\eqref{eqn:erosionModel}~\citep{qua-moo2018}, we set both
these parameters to one. The shear stress on $\gamma$ is
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
where, $\nn$ is the normal vector pointing into the body, and $\ss$ is
the unit tangent vector pointing in the counterclockwise direction. We
simulate erosion by alternating between solving the fluid equations and
advancing the eroding grains.  The strength of $\UU$ is adjusted at each
time step to achieve a constant pressure drop across the channel,
motivated by the geological situation of a porous medium connecting two
regions of  fixed hydraulic heads.

@@

Consider first an isotropic porous medium with {\em scalar} permeability $k$. Furthermore, we seek to characterize the medium by a single, bulk value permeability, so we assume that $k$ is spatially constant.
Equivalently, we can introduce a {\em resistivity} simply defined as the reciprocal of permeability, $r = 1/k$. Then taking the horizontal component gives
\begin{equation}
\label{q1}
q_1(x,y) = -\frac{1}{r} \pderiv{p}{x}(x,y)
\end{equation}
Consider the horizontal discharge, $q_1(x,y)$, averaged over a vertical slice
\begin{equation}
\qavg := \frac{1}{2} \int_{-1}^{1} q_1(x_0, y) dy
\end{equation}
By conservation of mass, $\qavg$ is independent of the horizontal position $x_0$. Since $\bq$ is a volume-average of the tracer velocity, $\qavg$ can be equivalently obtained by averaging the tracer velocity, which is how we will measure this quantity in the simulations.

@@

Integrating \eqref{q1} over the porous-medium domain $[-1, 1] \times [-1,1]$, applying the fundamental theorem of calculus, and rearranging gives
\begin{equation}
r = \left( \frac{\pup - \pdn}{2 \qavg} \right)
\end{equation}
This exact formula gives the resistivity based on measurements of the upstream/downstream pressure and the discharge.

@@

The resistivity is directly related to the drag exerted by the bodies. Indeed, another way to express the Stokes equations is $\grad \cdot \bvec{\sigma} = 0$ where $\bvec{\sigma} = -p \bvec{I} + \mu \left( \grad \bu + \grad \bu^T \right)$ is the stress tensor. The divergence theorem gives
\begin{equation}
\label{divthm}
0 = \int_{\Omega} \grad \cdot \bvec{\sigma} \, dV 
= \int_{\partial \Omega} \bvec{\sigma} \cdot \nhat \, dS
\end{equation}
for any domain $\Omega$.  Consider $\Omega$ to be inside $[-x_0, x_0] \times [-1, 1]$, where we will choose $x_0 > 1$ so that $\Omega$ includes the entire porous region.
The boundary integral on the right-hand-side of \eqref{divthm} has contributions from the boundaries of the erodible bodies $\gamma$ and from the outer boundary $\Gamma$, since $\partial \Omega = \gamma \cup \Gamma$. In particular, the total drag exerted by the erodible bodies is given by
\begin{equation}
\FD = (D_1, D_2) = -\int_{\gamma} \bvec{\sigma} \cdot \nhat \, dS 
\end{equation}
which includes both pressure and viscous drag. The minus sign is due to the orientation of the normal vector. We will mostly be interested in the horizontal component $D_1$.

Meanwhile, the horizontal component of the other term is
\begin{align}
\ex \cdot \int_{\Gamma} \bvec{\sigma} \cdot \nhat \, dS = 
& \int_{-1}^{1} p(-x_0, y) - p(x_0, y) - 2\mu (u_x(-x_0,y) - u_x(x_0,y)) dy \, + \\
& \int_{-x_0}^{x_0} \mu (u_y(x,1) - u_y(x,-1)) dx
\end{align}
If we take $x_0$ sufficiently far away, then the viscous terms involving $u_x$ will become small as the velocity field tends to a uniform flow. Meanwhile, the contribution from the pressure is well approximated by the upstream and downstream average pressures $\pup$ and $\pdn$ as introduced in the previous section. This gives 
\begin{equation}
D_1 \approx 2 (\pup - \pdn)
\end{equation}

Therefore we have the relationship
\begin{equation}
r = \frac{D_1}{4 \qavg}
\end{equation}

>>>>

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% COMMANDS
% Basic editing
\newcommand{\vsp}[1]{\vspace{#1 pc} \noindent}
\newcommand{\np}{\newpage \noindent}
\newcommand{\tocite}{ {\color{blue}(to cite)} }
\newcommand{\nick}[1]{ {\color{red} #1} }
% For real and imaginary, could use \Re or \Im, or \mathcal{R}, or \text{Re}
\newcommand{\Real}{\Re}
\newcommand{\Imag}{\Im}

%---------------------------------------------------%
% From JFM 2020
\newcommand{\bd}{{\partial}}
\newcommand{\bigO}{{\mathcal{O}}}
\newcommand{\cc}{{\mathbf{c}}}
\newcommand{\CC}{{\mathbb{C}}}
\newcommand{\DD}{{\mathcal{D}}}
\newcommand{\DDD}{{\boldsymbol{\mathcal D}}}
\newcommand{\eeta}{{\boldsymbol\eta}}
\newcommand{\ff}{{\mathbf{f}}}
\newcommand{\grad}{{\nabla}}
\newcommand{\II}{{\mathbf{I}}}
\newcommand{\iin}{\mathrm{in}}
\newcommand{\llambda}{{\boldsymbol\lambda}}
\newcommand{\nn}{{\mathbf{n}}}
\newcommand{\NN}{{\mathcal{N}}}
\newcommand{\out}{\mathrm{out}}
\newcommand{\rr}{{\mathbf{r}}}
\renewcommand{\Re}{{\operatorname{Re}}}
\renewcommand{\Im}{{\operatorname{Im}}}
\newcommand{\RR}{{\mathbb{R}}}
\renewcommand{\ss}{{\mathbf{s}}}
\newcommand{\tar}{\mathrm{tar}}
\newcommand{\bary}{\mathrm{bary}}
\newcommand{\trap}{\mathrm{trap}}
\newcommand{\uu}{{\mathbf{u}}}
\newcommand{\UU}{{\mathbf{U}}}
\newcommand{\vv}{{\mathbf{v}}}
\newcommand{\xx}{{\mathbf{x}}}
\newcommand{\xxi}{{\boldsymbol{\xi}}}
\newcommand{\yy}{{\mathbf{y}}}
\newcommand{\mcaption}[2]{\caption{\small \em #1}\label{#2}} 
\newcommand{\secref}[1]{\ref{#1}}
\def\gap{\hspace*{.2in}}
% Nick's below
\newcommand{\pderiv}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\ppd}[2]{\frac{\partial^2 #1}{{\partial #2}^2}}
\newcommand{\abs}[1]{\left| #1 \right|}
\newcommand{\Vn}{V_\nn}
\newcommand{\Vs}{V_\ss}
\newcommand{\CE}{C_E}
%---------------------------------------------------%

% From Nick's Latex Notes
\newcommand{\bvec}[1]{\mathbf{#1}}
%{\ensuremath{\boldsymbol{#1}}}
\newcommand {\bq} {\bvec{q}}
\newcommand{\mean}[1]{\left< #1 \right>}
\newcommand{\qavg}{\bar{q}}
\newcommand{\pavg}{\bar{p}}
\newcommand{\pup}{p_u}
\newcommand{\pdn}{p_d}
\newcommand{\stress}{{\boldsymbol \sigma}}
\newcommand{\FD}{\bvec{F}_d}
\newcommand{\ex}{ {\bvec{e}}_1}

@@

% NOTE: THIS SECTION HAS BEEN PLAGIARIZED FROM JFM 2020
% SO WE NEED TO PARAPHRASE IT
%\begin{comment}
%-----------------------------------------------------------------------------------------------%
\section{Governing Equations}
\label{sec:formulation}
%-----------------------------------------------------------------------------------------------%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\citep{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of the boundary of $[-3,3] \times [-1,1]$.  All eroding
bodies are placed in $[-1,1] \times [-1,1]$ to create a buffer region
that allows the flow profile imposed at the inlet to transition to the
more complex flow intervening between the bodies. Neglecting inertial
forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here $\uu$ is the fluid velocity, $p$ is the pressure, \nick{\st{$\UU$ is a
prescribed Hagen-Poiseuille velocity field}, $\UU$ will be far-field uniform flow $\UU = (U,0)$, where the value $U = U(t)$ is adjusted to keep the pressure drop constant in time,}
%
and $\Vn$ is the normal velocity of $\gamma$. Since the rate of erosion is much slower than the background flow, a quasi-steady approximation justifies the no-slip boundary condition. Parameters include the fluid viscosity $\mu$ and the material-dependent erosion constant $\CE$. By non-dimensionalizing equations~\eqref{eqn:erosionModel}~\citep{qua-moo2018}, we set both these parameters to one. The shear stress on $\gamma$ is
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
where, $\nn$ is the normal vector pointing into the body, and $\ss$ is
the unit tangent vector pointing in the counterclockwise direction. We
simulate erosion by alternating between solving the fluid equations and
advancing the eroding grains.  The strength of $\UU$ is adjusted at each
time step to achieve a constant pressure drop across the channel,
motivated by the geological situation of a porous medium connecting two
regions of  fixed hydraulic heads.

@@

Consider first an isotropic porous medium with {\em scalar} permeability $k$. Furthermore, we seek to characterize the medium by a single, bulk value permeability, so we assume that $k$ is spatially constant.
Equivalently, we can introduce a {\em resistivity} simply defined as the reciprocal of permeability, $r = 1/k$. Then taking the horizontal component gives
\begin{equation}
\label{q1}
q_1(x,y) = -\frac{1}{r} \pderiv{p}{x}(x,y)
\end{equation}
Consider the horizontal discharge, $q_1(x,y)$, averaged over a vertical slice
\begin{equation}
\qavg := \frac{1}{2} \int_{-1}^{1} q_1(x_0, y) dy = U
\end{equation}
By conservation of mass, $\qavg$ is independent of the horizontal position $x_0$ and is equal to the imposed value $U$. Since $\bq$ is a volume-average of the tracer velocity, $\qavg$ can be equivalently obtained by averaging the tracer velocity, which is how we will measure this quantity in the simulations.

@@

Integrating \eqref{q1} over the porous-medium domain $[-1, 1] \times [-1,1]$, applying the fundamental theorem of calculus, and rearranging gives
\begin{equation}
r = \frac{\pup - \pdn}{2 U}
\end{equation}
This exact formula gives the resistivity based on measurements of the upstream/downstream pressure and the discharge.

@@

The resistivity is directly related to the drag exerted by the bodies. Indeed, another way to express the Stokes equations is $\grad \cdot \stress = 0$ where $\stress = -p \bvec{I} + \mu \left( \grad \uu + \grad \uu^T \right)$ is the stress tensor. The divergence theorem gives
\begin{equation}
\label{divthm}
0 = \int_{\Omega} \grad \cdot \stress \, dV 
= \int_{\partial \Omega} \stress \cdot \nn \, ds
\end{equation}
for any domain $\Omega$.  Consider $\Omega$ to be inside $[-x_0, x_0] \times [-1, 1]$, where we will choose $x_0 > 1$ so that $\Omega$ includes the entire porous region.
The boundary integral on the right-hand-side of \eqref{divthm} has contributions from the boundaries of the erodible bodies $\gamma$ and from the outer boundary $\Gamma$, since $\partial \Omega = \gamma \cup \Gamma$. In particular, the total drag exerted by the erodible bodies is given by
\begin{equation}
\FD = -\int_{\gamma} \stress \cdot \nn \, ds
\end{equation}
which includes both pressure and viscous drag. The minus sign is due to the orientation of the normal vector. We will mostly be interested in the horizontal drag $\FD \cdot \ex$. \\

\nick{Need to check sign on drag and check whether I want the drag {\em on} the bodies or exerted {\em by} the bodies.} \\

Consider now the contribution to \eqref{divthm} from the outer boundary $\Gamma$. The horizontal component of this term is
\begin{align}
\ex \cdot \int_{\Gamma} \stress \cdot \nn \, ds = 
& \int_{-1}^{1} p(-x_0, y) - p(x_0, y) - 2\mu (u_x(-x_0,y) - u_x(x_0,y)) \, dy \, + \\
& \int_{-x_0}^{x_0} \mu (u_y(x,1) - u_y(x,-1)) dx
\end{align}
If we take $x_0$ sufficiently far away, then the viscous terms involving $u_x$ will become small as the velocity field tends to a uniform flow. Meanwhile, the contribution from the pressure is well approximated by the upstream and downstream average pressures $\pup$ and $\pdn$ as introduced in the previous section. This gives 
\begin{equation}
\FD \cdot \ex \approx 2 (\pup - \pdn)
\end{equation}

Therefore we have the relationship
\begin{equation}
r \approx \frac{1}{4 U}  \, \FD \cdot \ex
\end{equation}
